---
output:
  html_document: default
  pdf_document: default
---


# **NYPD Shooting Incident Data (Historic)**
Metadata Updated: November 10, 2020

List of every shooting incident that occurred in NYC going back to 2006 through the end of the previous calendar year.

This is a breakdown of every shooting incident that occurred in NYC going back to 2006 through the end of 
the previous calendar year. This data is manually extracted every quarter and reviewed by the Office of 
Management Analysis and Planning before being posted on the NYPD website. Each record represents a shooting 
incident in NYC and includes information about the event, the location and time of occurrence. In addition, 
information related to suspect and victim demographics is also included. 

  **<u>Source: </u>** Data.GOV
  
  **<u>Datafile location:</u>**  https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD
  
  **<u>Datafile Name:</u>**  NYPD_Shooting_Incident_Data__Historic_.csv


## Loading Necesary Packages and Data File:
 
 ... Define Package load Function
```{r load_pkgs, echo=FALSE, include=FALSE}
pkg_install <- function (Package_List) 
{   
    
    # convert arguments to vector
    packages <- Package_List
    
    # start loop to determine if each package is installed
    for(package in packages){
        
        if(package %in% (.packages()))
        {
            # message(package," is already loaded") 
        }
        else 
        {
            # if package is installed locally, load
            if(package %in% rownames(installed.packages()))
            {
                # message("Loading Package: ",package) 
                do.call('library', list(package))
            }
            # if package is not installed locally, download, then load
            else 
            {
                # message("Installing Package: ",package) 
                install.packages(package)
                # message("Loading Package: ",package) 
                do.call("library", list(package))
            }
        }
    } 
}

```
  ... Load packages
```{r Package_loading, echo=TRUE}
Packages_List_DS_Class <- c("psych", "stats", "lavaan", "tidyverse", "odbc", "lubridate", "Hmisc","summarytools","scales")
pkg_install(Packages_List_DS_Class)

```
 ... data importing: NYPD Shooting data and Borough population data 
```{r NYPD_Data, echo=TRUE}
NYPD_Shooting_Datafile_location <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
NYPD_Shooting_Data <- read.csv(NYPD_Shooting_Datafile_location)
```
 
  ... Check Data values in each field
```{r checking_data, echo=TRUE}
 describe(NYPD_Shooting_Data)
```

... change data types

 ```{r Modify_Classes}
 NYPD_Shooting_Data$INCIDENT_KEY <- as.character(NYPD_Shooting_Data$INCIDENT_KEY)
 NYPD_Shooting_Data$PRECINCT <- as.character(NYPD_Shooting_Data$PRECINCT)
 NYPD_Shooting_Data$JURISDICTION_CODE <- as.character(NYPD_Shooting_Data$JURISDICTION_CODE) 
 ```
 
 ... change appropriate fields to factors for Shooting data an Population data
 
```{r Fields_to_factors, echo=TRUE}
Factor_Fields_shooting_Data <- c("BORO","PRECINCT","JURISDICTION_CODE","STATISTICAL_MURDER_FLAG","VIC_AGE_GROUP","VIC_SEX","VIC_RACE","PERP_RACE","PERP_SEX","PERP_AGE_GROUP")
NYPD_Shooting_Data[,Factor_Fields_shooting_Data] <- lapply(NYPD_Shooting_Data[,Factor_Fields_shooting_Data] , factor) 


```
 
 ... Remove fields due to missing data for shooting data: LOCATION_DESC, PERP_AGE_GROUP, PERP_SEX
```{r Remove_Fields, echo=FALSE}
NYPD_Fields_To_Remove <- c("LOCATION_DESC", "PERP_AGE_GROUP", "PERP_SEX")
NYPD_Shooting_Data_Subset <- NYPD_Shooting_Data[ , ! names(NYPD_Shooting_Data) %in% NYPD_Fields_To_Remove] 

```

 ... rename fields for data joining and readability
 
```{r rename_fields, echo=FALSE}
NYPD_Shooting_Data <- rename(NYPD_Shooting_Data, Borough_Name = BORO)
NYPD_Shooting_Data_Subset <- rename(NYPD_Shooting_Data_Subset, Borough_Name = BORO)
``` 


## **Analysis by Victim race of Black/African American**



 ... stacked bar chart Victim by race and borough
 
```{r Victim_by_race}
ggplot(NYPD_Shooting_Data_Subset, aes(x=VIC_RACE, fill=Borough_Name)) + geom_bar() + theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5)) + ggtitle("Victim by Race and Borough")
```


**Percentage of Victim totals by Race**

```{r cal_vic_race_totals,echo=TRUE}
VIC_Race_Percentages <- NYPD_Shooting_Data_Subset  %>% group_by(VIC_RACE) %>% summarise(Count = length(VIC_RACE),Total = sum(length(NYPD_Shooting_Data_Subset$VIC_RACE), na.rm = TRUE),Percent = (length(VIC_RACE)) / sum(length(NYPD_Shooting_Data_Subset$VIC_RACE), na.rm = TRUE))
VIC_Race_Percentages$Percentages <- percent(VIC_Race_Percentages$Percent) 
knitr::kable(VIC_Race_Percentages, align = 'c')
```

## Initial Findings

We notice a rather large percentage of the shootings involve victims that have a racial value of 
Black or African American and many of the shootings occurring in Brooklyn.

A review of the population breakdown might help see if this trend is due to the racial make
up of the boroughs themselves

A good downloadable dataset by borough was not readily available in a friendly import format or had 
issues on data accuracy depending on the source due to limited funding scopes as noted on the New York data website

The data with population and percentage breakdowns have been extracted from the file found here:

https://www1.nyc.gov/assets/planning/download/office/data-maps/nyc-population/acs/demo_2016acs5yr_nyc.xlsx

This is 2016 data and show the total black population for each Borough with the percentage that makes up of the 
total population there in that borough

**Black or African American Population break down by total and Percentage in each Bourough and New York itself**

Borough        |  Population        |  Percent Total
-------------- | ------------------ | ----------------		
Bronx 	       | Estimate	476,525   |  Percent	33.2%
Brooklyn	     | Estimate	871,223   |  Percent	33.4%
Manhattan	     | Estimate	244,787   |  Percent	15.0%
Queens 	       | Estimate	423,451   |  Percent	18.3%
Staten Island  | Estimate	48,941    |  Percent	10.3%
New York city	 | Estimate	2,064,927 |  Percent	24.4%

## Further Analysis

While the percentage of black/African American victims was at 71%, none of the Boroughs
show a population near that high to warrant the victim racial identity of Black/African American being
so high.

Further review below shows more graphical breakdowns of Victim by Age group and Gender with comparable 
racial graph of the Perpetrator's race which appears to mirror the victim's race.

The perpetrator's racial data though suffers from a large amount of it being empty and does not 
give use good feel if racially motivated violence is going on or not. 

Looks like we need better view by Borough from a race perspective


... **Graphs**
 
 ... Victim by Age Group
 
```{r Victim_by_Age_Group, echo=FALSE}
ggplot(NYPD_Shooting_Data_Subset, aes(x=VIC_AGE_GROUP, fill=Borough_Name)) + geom_bar() + theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5)) + ggtitle("Victim by Age and Borough")
```
 
 

  ... Victim by Gender 
  
```{r Victim_by_Gender, echo=FALSE}
ggplot(NYPD_Shooting_Data_Subset, aes(x=VIC_SEX, fill=Borough_Name)) + geom_bar() + theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5)) + ggtitle("Victim by Gender and Borough")
```
 
 

  ... Victim by Perp Race 
  
```{r Victim_by_Perp_Race, echo=FALSE}
ggplot(NYPD_Shooting_Data_Subset, aes(x=PERP_RACE, fill=Borough_Name)) + geom_bar() + theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))
```

The 3 previous graphs kind of lead us in circle and still show that race seems to be an issue.
We might be better served by again taking population in account.


```{r Black_Victim_Race_Total, echo=TRUE}
 VIC_Race_Black_Total <- NYPD_Shooting_Data_Subset  %>% filter(VIC_RACE %in% c("BLACK HISPANIC","BLACK")) %>% summarise(Count = length(VIC_RACE))
```
```{r Black_Vick_Race_Total, echo=TRUE}
VIC_Race_Black_Total <- NYPD_Shooting_Data_Subset  %>% filter(VIC_RACE %in% c("BLACK HISPANIC","BLACK")) %>% summarise(Count = length(VIC_RACE))  
```
```{r Create_NY_Pop_Set, echo=TRUE}
NY_Black_Pop = data.frame(
     Borough_Name = c("Bronx", "Brooklyn", "Manhattan", "Queens","Staten Island"),
     Black_Population = c(476525, 871223, 244787,423451,48941)
)
NY_Black_Pop_Total <- sum(NY_Black_Pop$Black_Population)
```
```{r show_Black_NY_pop, echo=TRUE}
Black_Race_percent <- (VIC_Race_Black_Total/NY_Black_Pop_Total)
```

**Black Victim Total:** `r prettyNum(VIC_Race_Black_Total, big.mark = ",", scientific = FALSE)`

**Black population Total:** `r prettyNum(NY_Black_Pop_Total, big.mark = ",", scientific = FALSE)`

**Black Victim Race as percent of Black Population:** `r percent(Black_Race_percent$Count, accuracy = 0.3)`



## Conclusion

  The data does leave several things open for questioning:
  
   1) While the victim rate for Black/African Americans is high, another breakdown by 
      victims per incident may be useful to denote how often these are isolated incidents
      versus a gang related incident maybe involving a rapid fire type of weapon.

   2) No indication in the data of what type of incident it was. Usually there is an initial call
      into police for domestic violence, burglary in process...etc. There is a location field
      but, it is missing a lot of the times does not give us any insight into the incident itself.
      
   3) Bias: This limited overview might lead one to believe it's a just a black on black violence thing. 
      But, would be a definite problem due to it being biased report that focused just on the victim's
      race being Black/African American and blamed the color of their skin for reason behind the shootings. 

   4) The final comparison of Black Victim's race total to the Black population in NY brings better
      perspective to the high Percentage breakouts when just looking at race in this set only.
      One's view becomes skewed and misses part of the bigger picture
      
   5) More socioeconomic data could be incorporated to highlight disparities between Boroughs. Many 
      times crime is tied to areas that have a lower economic status rather than race being the issue.
      This would help focus on actual problems leading to the shootings and help mitigate the bias in
      it's limited focus.
      